---
title: "Pop-gen Sample Selection"
author: "Holden"
date: "2025-07-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/holdenfox/Desktop/shrike-gen/map")
```

Howdy folks!

Often in population genomics projects you end up with more genetic samples than you can actually sequence. And it can actually be kinda tricky, or at least time-consuming, to decide which samples to include on a sequencing lane and which ones to drop. I struggle with this a bunch, especially when working with data from many different sources, somewhat randomly sampled, across large time-scales, and including various tissue types and dna quality. But it is important to make smart choices for what to sequence, so that you maximize the potential use of the genomic data generated. That's why I am creating a set of tools to provide a robust system for ranking genomic samples based on their geographic distance, sample type (associated w/ dna quality), and age. The goal here is to prioritize and automate selecting samples that can form viable clusters for analysis of population structure, genotype-environment associations, and more.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(geosphere)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(plotly)
```

Ok so the primary criteria for ranking samples probably ought to be based on geographic distance. So let's start with that. Whatever the state of your metadata, you should probably have a column with the latitude and longitude of each sample.

I'm going to use a combined metadata sheet for shrikes for testing. The important information that should be complete are the cooridnates (lat, long), sample type (I have a variety of sample types to choose from), and an ID.
```{r, message=FALSE, warning=FALSE}
meta <- read_csv("../complete-winter-shrike-metadata/winter_metadata_feathers_and_blood.csv")

meta <- meta %>% filter(!is.na(Lat) & !is.na(Long))

# let's get rid of the Susan Heath wintering shrike samples that all but one turned out to be residents.
meta <- meta %>% filter(Contact != "Susan Heath")

# extract coordinates
coords <- meta %>% dplyr::select(Long, Lat)

# calculate distance matrix using Haversine distance
dist_matrix <- distm(as.matrix(coords), fun = distHaversine)

# clustering
hclust_groups <- hclust(as.dist(dist_matrix), method = "complete") # complete method here avoids chaining, ensuring all samples in a cluster are within the specified distance of each other. If you want some flexibility you could try "average"

# cut tree at 111.111 km to assign spatial clusters
meta$GroupID <- cutree(hclust_groups, h = 111111)

# create population labels and filter for minimum cluster size
meta <- meta %>%
  mutate(Pop = paste0("Cluster_", GroupID)) %>%
  group_by(Pop) %>%
  filter(n() > 3) %>%  # keep clusters with ≥4 individuals
  ungroup()

# cluster size distribution
cluster_sizes <- meta %>%
  group_by(Pop) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n))

print(cluster_sizes)
```
```{r, message=FALSE, warning=FALSE, echo=FALSE}
# summary statistics
cat("Number of clusters with ≥4 individuals:", length(unique(meta$Pop)), "\n")
cat("Total individuals retained:", nrow(meta), "\n")
```

Map it so we can visualize where the samples are
```{r, message=FALSE, warning=FALSE}
# import base map
map <- ne_states(country = c("United States of America", "Mexico"), returnclass = 'sf')

# import shrike range from ebird
shrike_range <- st_read("/Users/holdenfox/Desktop/shrike-gen/meta/logshr_range_2022/logshr_range_2022.gpkg") %>%
  filter(season == "nonbreeding") %>%
  st_set_crs(4326)

# create colors for clusters
n_clusters <- length(unique(meta$Pop))
cluster_colors <- rainbow(n_clusters)
names(cluster_colors) <- unique(meta$Pop)

meta <- meta %>%
  group_by(Pop) %>%
  mutate(cluster_size = n()) %>%
  ungroup()

# plot it
p <- ggplot() +
  geom_sf(data = map, fill = "white", color = "black", size = 0.1) +
  geom_sf(data = shrike_range, fill = "grey", color = NA, alpha = 0.5) +
  geom_point(data = meta, aes(x = Long, 
                              y = Lat, 
                              color = Pop,
                              text = paste0(
                                "ID: ", SampleID, "<br>",
                                "Cluster: ", Pop, "<br>",
                                "Cluster Size: ", cluster_size, "<br>",
                                "Sequenced: ", Sequenced, "<br>",
                                "SampleType: ", SampleType, "<br>",
                                "CollectionDate: ", CollectionDate, "<br>")), 
             size = 2, position = position_jitter(width = 0.1, height = 0.1)) +
  scale_colour_manual(values = cluster_colors, name = "Spatial Clusters", guide = "none") +
  coord_sf(xlim = c(-140, -65), ylim = c(15, 50)) + 
  theme_minimal() +
  labs(title = "All clusters with ≥4 individuals",
       x = "Longitude",
       y = "Latitude")


ggplotly(p, tooltip = "text")
```

I've been referencing the map and filtering out clusters to dedensify certain parts of the sample distribution in order to increase the sample number per cluster given a strict budget. I've kinda been doing this manually. That takes place here.
```{r, message=FALSE, warning=FALSE}
meta <- left_join(meta, cluster_sizes, by = "Pop") %>% 
  filter(!Pop %in% c("Cluster_18", "Cluster_31", "Cluster_34", "Cluster_36", "Cluster_37", "Cluster_51", "Cluster_61","Cluster_70", "Cluster_14", "Cluster_32", "Cluster_35", "Cluster_33", "Cluster_15", "Cluster_52"))
```

Here will be a function to pick the number of samples you should select based on the number of lanes you have available and the target depth you want to achieve. Holden use your post rmdup coverage to fill this in.
```{r, message=FALSE, warning=FALSE}

```

Ok so now the ranking and selection portion...
```{r, message=FALSE, warning=FALSE}
# Step 1: Define sample type ranking hierarchy
# give lower quality samples a lower rank
sample_type_ranking <- data.frame(
  SampleType = c("Blood", "Blood, Feather", "DNA", "DNA, Feather", "Feather", "Toe pad"),
  type_rank = c(1, 1, 2, 3, 4, 5),  # 1 = highest priority (Blood and Blood,Feather treated same)
  stringsAsFactors = FALSE
)

# Step 2: Determine the age of the sample based on the collection date. Newer samples will get a higher priority.
# This should be able to handle multiple Date formats
calculate_age <- function(collection_date, reference_date = Sys.Date()) {
  if (is.na(collection_date) || collection_date == "") return(NA)
  
  # convert to date if it's a string
  if (is.character(collection_date)) {
    # try a couple of different formats
    parsed_date <- tryCatch({
      # M/D/YYYY
      as.Date(collection_date, format = "%m/%d/%Y")
    }, error = function(e) {
      tryCatch({
        # M/D/YY
        as.Date(collection_date, format = "%m/%d/%y")
      }, error = function(e) {
        tryCatch({
          as.Date(collection_date)
        }, error = function(e) {
          return(NA)
        })
      })
    })
    
    collection_date <- parsed_date
  }
  
  if (is.na(collection_date)) return(NA)
  
  # get the age in years
  age_years <- as.numeric(difftime(reference_date, collection_date, units = "days")) / 365.25
  return(age_years)
}

# Step 3: Rank samples, give already sequenced samples the highest priority
rank_samples <- function(meta, 
                        sample_type_col = "SampleType", 
                        date_col = "CollectionDate",
                        sequenced_col = "Sequenced",
                        prefer_recent = TRUE) {
  
  # add sample type ranking
  meta <- meta %>%
    left_join(sample_type_ranking, by = setNames("SampleType", sample_type_col)) %>%
    mutate(
      # if sample type not found in ranking, assign lowest priority maybe change this, b/c atm YOU MUST HAVE SAMPLE TYPE INFORMATION TO RANK AND EFFECTIVELY CHOOSE SAMPLES
      type_rank = ifelse(is.na(type_rank), max(sample_type_ranking$type_rank) + 1, type_rank)
    )
  
  # rank by age
  meta <- meta %>%
    mutate(
      age_years = sapply(get(date_col), calculate_age),
      # prefer recent samples 
      age_rank = if (prefer_recent) {
        rank(age_years, ties.method = "min", na.last = TRUE)
      } else {
        rank(-age_years, ties.method = "min", na.last = FALSE)
      }
    )
  
  # within each cluster, 
  meta <- meta %>%
    group_by(Pop) %>%
    mutate(
      # check if already sequenced
      is_sequenced = ifelse(is.na(get(sequenced_col)), FALSE, 
                           toupper(get(sequenced_col)) == "Y"),
      
      # create composite rank
      # already sequenced samples get rank 0-999, others get 1000+
      composite_rank = ifelse(is_sequenced, 
                             age_rank,  # just age rank for sequenced samples (0-999)
                             type_rank * 1000 + age_rank),  # type + age for unsequenced (1000+)
      
      cluster_rank = rank(composite_rank, ties.method = "min")
    ) %>%
    ungroup()
  
  return(meta)
}

# Step 4: Actually select samples
select_samples <- function(ranked_meta, 
                           max_samples_per_cluster = 8,
                           min_samples_per_cluster = 4,
                           total_budget = NULL) {
  
  # always select all sequenced samples firt
  sequenced_samples <- ranked_meta %>%
    filter(is_sequenced == TRUE)
  
  cat("Selected", nrow(sequenced_samples), "\n")
  
  # calculate the remaining budget for unsequenced samples
  remaining_budget <- if (!is.null(total_budget)) {
    max(0, total_budget - nrow(sequenced_samples))
  } else {
    NULL
  }
  
  if (!is.null(total_budget)) {
    cat("Remaining budget:", remaining_budget, "\n")
  }
  
  # select unsequenced samples between min and max allowed
  unsequenced_candidates <- ranked_meta %>%
    filter(is_sequenced == FALSE)
  
  if (nrow(unsequenced_candidates) == 0) {
    cat("No unsequenced samples available\n")
    return(sequenced_samples)
  }
  
  # initial selection of unsequenced samples by cluster
  unsequenced_selected <- unsequenced_candidates %>%
    group_by(Pop) %>%
    arrange(cluster_rank) %>%
    slice_head(n = max_samples_per_cluster) %>%
    ungroup()
  
  # account for sequenced samples in max_samples_per_cluster limit
  sequenced_per_cluster <- sequenced_samples %>%
    count(Pop, name = "sequenced_count")
  
  unsequenced_selected <- unsequenced_selected %>%
    left_join(sequenced_per_cluster, by = "Pop") %>%
    mutate(sequenced_count = replace_na(sequenced_count, 0)) %>%
    group_by(Pop) %>%
    mutate(row_num = row_number()) %>%
    # keep only rows where row number <= allowed unsequenced samples for this cluster
    filter(row_num <= pmax(0, max_samples_per_cluster - sequenced_count)) %>%
    select(-sequenced_count, -row_num) %>%
    ungroup()
  
  # combine sequenced and unsequenced before checking min cluster sizes
  all_selected <- bind_rows(sequenced_samples, unsequenced_selected)
  
  # check ifmin cluster size is met
  cluster_sizes <- all_selected %>%
    count(Pop, name = "selected_count")
  
  clusters_below_min <- cluster_sizes %>%
    filter(selected_count < min_samples_per_cluster) %>%
    pull(Pop)
  
  if (length(clusters_below_min) > 0) {
    cat("Removing", length(clusters_below_min), "clusters with fewer than", min_samples_per_cluster, "samples:\n")
    print(clusters_below_min)
    
    # remove unsequenced samples from clusters below the min
    # sequenced samples are always kept
    unsequenced_selected <- unsequenced_selected %>%
      filter(!Pop %in% clusters_below_min)
    
    # update combined selection
    all_selected <- bind_rows(sequenced_samples, unsequenced_selected)
  }
  
  # apply budget constraint to unsequenced samples only
  if (!is.null(remaining_budget) && nrow(unsequenced_selected) > remaining_budget) {
    
    samples_to_remove <- nrow(unsequenced_selected) - remaining_budget
    cat("Need to remove", samples_to_remove, "unsequenced samples\n")
    
    removed_count <- 0
    
    while (removed_count < samples_to_remove && nrow(unsequenced_selected) > 0) {
      
      # count current samples per cluster
      current_cluster_sizes <- bind_rows(sequenced_samples, unsequenced_selected) %>%
        count(Pop, name = "total_count")
      
      # find cluster with most total samples that can lose unsequenced samples
      clusters_with_unsequenced <- unsequenced_selected %>%
        count(Pop, name = "unsequenced_count") %>%
        left_join(current_cluster_sizes, by = "Pop") %>%
        # only remove if cluster will still meet minimum after removal
        filter(total_count > min_samples_per_cluster) %>%
        arrange(desc(total_count)) %>%
        slice_head(n = 1) %>%
        pull(Pop)
      
      if (length(clusters_with_unsequenced) == 0) {
        cat("Unable to remove more unsequenced samples while maintaining minimum cluster size\n")
        break
      }
      
      target_cluster <- clusters_with_unsequenced[1]
      
      # remove the worst ranked unsequenced sample
      sample_to_remove <- unsequenced_selected %>%
        filter(Pop == target_cluster) %>%
        arrange(desc(cluster_rank)) %>%
        slice_head(n = 1)
      
      unsequenced_selected <- unsequenced_selected %>%
        filter(SampleID != sample_to_remove$SampleID)
      
      removed_count <- removed_count + 1
    }
    
    cat("Removed", removed_count, "unsequenced samples\n")
  }
  
  # combine sequenced + selected unsequenced
  selected_samples <- bind_rows(sequenced_samples, unsequenced_selected)

  
  return(selected_samples)
}



### USE RANKING AND SELCTION FUNCTION HERE

# Notes 
# 40 samples have already been sequenced, so number to select is total_budget - 40. I don't have the metadata for the samples Amanda and I collected, there are like 4 in Colorado and like 7 in Utah i believe? We should include all the CO ones and at leat 5 of the Utah ones. Doesn't matter which as long as they cluster together...

# apply ranking
ranked_meta <- rank_samples(meta, 
                           sample_type_col = "SampleType",  
                           date_col = "CollectionDate",     
                           sequenced_col = "Sequenced",     
                           prefer_recent = TRUE)

# select samples
selected_samples <- select_samples(ranked_meta, 
                                  max_samples_per_cluster = 8, # max number of samples per cluster
                                  min_samples_per_cluster = 4, # min number of samples per cluster
                                  total_budget = 230)  # total number of samples to select

write_csv(selected_samples, "LOSH_Winter_Library_Selection_JUL25.csv")

# selected samples including already sequenced by cluster
cluster_summary <- selected_samples %>%
  group_by(Pop) %>%
  summarise(
    selected_count = n(),
    available_count = first(n()),
    .groups = "drop"
  ) %>%
  arrange(desc(selected_count))

print(cluster_summary)

```

Visualize it
```{r, message=FALSE, warning=FALSE}
selected_samples <- selected_samples %>%
  group_by(Pop) %>%
  mutate(cluster_size = n()) %>%
  ungroup()

# plot it
p <- ggplot() +
  geom_sf(data = map, fill = "white", color = "black", size = 0.2) +
  geom_sf(data = shrike_range, fill = "grey", color = NA, alpha = 0.5) +
  geom_point(data = selected_samples, aes(x = Long, 
                              y = Lat, 
                              color = Pop,
                              text = paste0(
                                "ID: ", SampleID, "<br>",
                                "Cluster: ", Pop, "<br>",
                                "Cluster Size: ", cluster_size, "<br>",
                                "Sequenced: ", Sequenced, "<br>",
                                "CollectionDate: ", CollectionDate, "<br>")), 
             size = 2, position = position_jitter(width = 0.3, height = 0.3)) +
  scale_colour_manual(values = cluster_colors, name = "Spatial Clusters", guide = "none") +
  coord_sf(xlim = c(-140, -65), ylim = c(15, 50)) +  # Using your original study area bounds
  theme_minimal() +
  labs(title = "Samples Selected For Sequencing",
       x = "Longitude",
       y = "Latitude")


ggplotly(p, tooltip = "text")
```


###############################################################################################################

###alternate selection function

This doesn't work correctly yet.
```{r, message=FALSE, warning=FALSE, include = FALSE}
select_samples <- function(ranked_meta, 
                           max_samples_per_cluster = 8,
                           min_samples_per_cluster = 4,
                           total_budget = NULL) {
  
  
  sequenced_samples <- ranked_meta %>%
    filter(is_sequenced == TRUE)
  
  cat("Selected", nrow(sequenced_samples), "sequenced samples (always included)\n")
  
  
  remaining_budget <- if (!is.null(total_budget)) {
    max(0, total_budget - nrow(sequenced_samples))
  } else {
    NULL
  }
  
  if (!is.null(total_budget)) {
    cat("Remaining budget for unsequenced samples:", remaining_budget, "\n")
  }
  
  
  unsequenced_candidates <- ranked_meta %>%
    filter(is_sequenced == FALSE)
  
  if (nrow(unsequenced_candidates) == 0) {
    cat("No unsequenced samples available\n")
    return(sequenced_samples)
  }
  
  
  sequenced_per_cluster <- sequenced_samples %>%
    count(Pop, name = "sequenced_count")
  
  unsequenced_per_cluster <- unsequenced_candidates %>%
    count(Pop, name = "unsequenced_count")
  
  cluster_potential <- unsequenced_per_cluster %>%
    left_join(sequenced_per_cluster, by = "Pop") %>%
    mutate(sequenced_count = replace_na(sequenced_count, 0)) %>%
    mutate(total_potential = sequenced_count + unsequenced_count) %>%
    filter(total_potential >= min_samples_per_cluster)
  
  valid_clusters <- cluster_potential$Pop
  cat("Keeping", length(valid_clusters), "clusters that can meet minimum size requirement\n")
  
  
  unsequenced_candidates <- unsequenced_candidates %>%
    filter(Pop %in% valid_clusters)
  
  if (nrow(unsequenced_candidates) == 0) {
    cat("No valid unsequenced candidates after cluster filtering\n")
    return(sequenced_samples)
  }
  
  
  selected_unsequenced <- data.frame()
  
  
  current_per_cluster <- sequenced_samples %>%
    filter(Pop %in% valid_clusters) %>%
    count(Pop, name = "current_count")
  
  
  missing_clusters <- setdiff(valid_clusters, current_per_cluster$Pop)
  if (length(missing_clusters) > 0) {
    current_per_cluster <- bind_rows(
      current_per_cluster,
      data.frame(Pop = missing_clusters, current_count = 0)
    )
  }
  
  
  samples_selected <- 0
  max_iterations <- if (!is.null(remaining_budget)) remaining_budget else nrow(unsequenced_candidates)
  
  while (samples_selected < max_iterations && nrow(unsequenced_candidates) > 0) {
    
    
    available_clusters <- current_per_cluster %>%
      filter(current_count < max_samples_per_cluster) %>%
      pull(Pop)
    
    if (length(available_clusters) == 0) {
      cat("All clusters have reached maximum sample limit\n")
      break
    }
    
    
    best_sample <- unsequenced_candidates %>%
      filter(Pop %in% available_clusters) %>%
      arrange(cluster_rank) %>%
      slice_head(n = 1)
    
    if (nrow(best_sample) == 0) break
    
    
    selected_unsequenced <- bind_rows(selected_unsequenced, best_sample)
    
    
    cluster_pop <- best_sample$Pop
    current_per_cluster <- current_per_cluster %>%
      mutate(current_count = ifelse(Pop == cluster_pop, current_count + 1, current_count))
    
    
    unsequenced_candidates <- unsequenced_candidates %>%
      filter(SampleID != best_sample$SampleID)
    
    samples_selected <- samples_selected + 1
    
    if (samples_selected %% 50 == 0) {
      cat("Selected", samples_selected, "unsequenced samples\n")
    }
  }
  
  
  selected_samples <- bind_rows(sequenced_samples, selected_unsequenced)
  
  return(selected_samples)
}

```
